Investigating soundscapes perception through acoustic scenes simulation

### 2018-06-27 lafaySimulatedScenePerception.tex
done with reformating citations

One of the main goals of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}. Establishing a link between an environment (urban sound scene) and the induced human sensation (calmness) is equivalent to instantiating the mental representation of a specific sound scene (a calm urban sound scene). Several means have been considered in the literature to perform this task.

@@

First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. Working with actual stimuli makes it possible to analyze them, thus allowing the experimenter to gather a coarse-grained physical description of them.

@@

Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}. A large amount of quantitative and semantic information is then collected about the subject's representation of this type of sound environment. Unfortunately, without any reference to sound data, this representation can hardly be characterized physically.

@@

We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.

@@

To the best of our knowledge, only Bruce~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics. A lack of diversity in terms of sound sources is nonetheless mentioned by the authors that limit the strength of the outcomes given in this study.

@@

\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{Web Audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{Web Audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.

@@

\begin{itemize}
\item {sound events}: isolated sounds, precisely located in time, whose acoustical characteristics may change with respect to time;
\item {sound textures}: isolated sounds of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

@@

According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment \cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}, while events that cannot be isolated are grouped into a same stream.

%Maffiolo \cite{maffiolo_caracterisation_1999} met en évidence deux processus de catégorisation distincts, engagés en fonction de la capacité de l'auditeur à identifier des événements sonores. Elle montre l'existence de deux catégories cognitives abstraites d'environnements sonores respectivement appelées: «~les séquences événementielles~» et «~les séquences amorphes~».

Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes two separated categorization processes, either of which is triggered depending on the listener's ability to identify sound events. She shows that those processes lead to two abstract cognitive categories respectively termed "event sequences" and "amorphous sequences".

%Les séquences événementielles sont des environnements composés d'événements saillants et identifiables (\emph{démarrage de voiture}, \emph{voix d'homme}). Elles profitent d'une analyse descriptive basée sur l'identification des sources sonores. Les séquences amorphes sont des environnements dont il est difficile d'isoler des éléments distincts («~brouhaha de rue~», «~brouhaha de trafic~»). Elles profitent, elles, d'une analyse holistique, effectuée à partir d'indicateurs acoustiques (objectifs) globaux. %Cette différence de traitement entre scènes événementielles et scènes amorphes est relevée également dans \cite{guastavino2006ideal,dubois2006cognitive}.

@@

Concerning sound textures, \ie~sounds that have stable characteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound \cite{nelken2013ear}.

@@

In the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli \cite{nelken2013ear,mcdermott2013summary}. We thus consider in this study the soundscape as a ``\,skeleton of events on a bed of textures\,'' as coined in \cite{nelken2013ear}.

@@

The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure similar to the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources. For classes with a high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.

@@

Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

Each of those classes has a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle. The subject can browse the database by listening to those prototype sounds. The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.

@@

%La scène sonore est ainsi vue comme une somme de sources sonores, ou, si l'on admet la distinction opérée entre événements et textures sonores, «~un squelette d'événements sur un lit de textures~» \cite{nelken2013ear}. Cet outil de simulation est présenté en détail dans \cite{rossignol2015simscene}.

As such, the underlying model of the scene is a sum of sound sources. Further details about the simulation interface can be found in \cite{rossignol2015simscene}.

@@

The design of this experiment has been validated with a pilot study described in \cite{lafay2014new}.

@@

\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni). To do so, a V-test is considered \citep{rakotomalala2008tvpercent}. With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:

@@

It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non ideal scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie~the events.

@@

We observe a noticeable difference in terms of class choices between the i- and non ideal scenes. The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non ideal scenes. These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.

@@

Nevertheless, we notice some differences with \cite{guastavino2006ideal}: Guastavino’s results show that sounds of \emph{public transportation} are specific of ideal urban soundscapes. The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors. Thus, in our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.

@@

This difference may be explained by the nature of the experimental protocol used in the two studies. As in our study, Guastavino asks the subjects to describe an environment. She asks them to perform this task starting only from their memories, whereas in our case subjects perform the same task using actual sound samples that they can listen to. The fact that subjects in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact. Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment \cite{lavandier2006contribution}.

@@

Considering $E$ and $T$ separately, it appears that: 1) the rank 5-precision obtained with $E$ is similar to the one obtained with $ET$ ; 2) the precision obtained with $T$ is always lower than the one obtained with $E$, by $10\%$ to $15\%$. Those results indicate that the semantic information that is discriminative is mostly carried by the events. Those results are in line with \cite{maffiolo_caracterisation_1999}. As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie~by identifying the sources.

@@

The fact that the pleasantness of the ideal scenes is not correlated to global physical features, contrary to the pleasantness of non ideal scenes, has also been studied recently \cite{gozalo2015relationship}.

@@

This strongly confirms that ideal markers do have a positive impact on the perception of an environment. The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should be possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}. Those conclusions are in line with the positive approach introduced by Schafer in \cite{schafer1977tuning}.

@@

Considering sound scene synthesis or composition as part of an experimental protocol requires the availability of software resources that are easier and easier to master. While most processing framework that have been considered are developed in native language for efficiency purposes like Marsyas \cite{tzanetakis2000marsyas} or Clam \cite{amatriain2006clam}, many advanced sound manipulation.

Closer to our sound model is the Tapestrea framework \cite{misra2007musical} that allows the manipulation sounds trough wavelet and sinusoidal modeling to allow sound manipulations for doing acousmatic composition or sound
design. The main issue with considering this framework for soundscape perception studies is that the sounds are processed using sound models that would strongly decrease the ecological validity of the study. That being said, the sequencing model is quite close to the one considered in this study.

>>>>

As asserted by~\cite{aletta2016soundscape}, one of the main goals of soundscape studies is to identify which components of the soundscape influence human perception. Establishing a link between an environment (urban sound scene) and the induced human sensation (calmness) is equivalent to instantiating the mental representation of a specific sound scene (a calm urban sound scene). Several means have been considered in the literature to perform this task.

@@

First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness), see~\cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. Working with actual stimuli makes it possible to analyze them, thus allowing the experimenter to gather a coarse-grained physical description of them.

@@

Second, a subject can be asked to describe a given sound environment, see~\cite{guastavino2006ideal, dubois2006cognitive}. A large amount of quantitative and semantic information is then collected about the subject's representation of this type of sound environment. Unfortunately, without any reference to sound data, this representation can hardly be characterized physically.

@@

We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene, see~\cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene, as found by~\cite{gozalo2015relationship,ricciardi2015sound}.

@@

To the best of our knowledge, only~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics. A lack of diversity in terms of sound sources is nonetheless mentioned by the authors that limit the strength of the outcomes given in this study.

@@

\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{Web Audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{Web Audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} visualization library proposed by~\cite{d32011}.

@@

The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts written by~\cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.

@@

\begin{itemize}
\item {sound events}: isolated sounds, precisely located in time, whose acoustical characteristics may change with respect to time;
\item {sound textures}: isolated sounds of long duration whose acoustical characteristics are stable with respect to time, as expressed by~\cite{saint1995classification}.
\end{itemize}

@@

According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment~\cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}, while events that cannot be isolated are grouped into a same stream.

%Maffiolo~\cite{maffiolo_caracterisation_1999} met en évidence deux processus de catégorisation distincts, engagés en fonction de la capacité de l'auditeur à identifier des événements sonores. Elle montre l'existence de deux catégories cognitives abstraites d'environnements sonores respectivement appelées: «~les séquences événementielles~» et «~les séquences amorphes~».

Maffiolo~\cite{maffiolo_caracterisation_1999} distinguishes two separated categorization processes, either of which is triggered depending on the listener's ability to identify sound events. She shows that those processes lead to two abstract cognitive categories respectively termed "event sequences" and "amorphous sequences".

%Les séquences événementielles sont des environnements composés d'événements saillants et identifiables (\emph{démarrage de voiture}, \emph{voix d'homme}). Elles profitent d'une analyse descriptive basée sur l'identification des sources sonores. Les séquences amorphes sont des environnements dont il est difficile d'isoler des éléments distincts («~brouhaha de rue~», «~brouhaha de trafic~»). Elles profitent, elles, d'une analyse holistique, effectuée à partir d'indicateurs acoustiques (objectifs) globaux. %Cette différence de traitement entre scènes événementielles et scènes amorphes est relevée également dans~\cite{guastavino2006ideal,dubois2006cognitive}.

@@

Concerning sound textures, \ie~sounds that have stable characteristics over time,~\cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound~\cite{nelken2013ear}.

@@

In the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli~\cite{nelken2013ear,mcdermott2013summary}. We thus consider in this study the soundscape as a ``\,skeleton of events on a bed of textures\,'' as coined in~\cite{nelken2013ear}.

@@

The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure similar to the vertical axis of the categorical organization proposed by~\cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources. For classes with a high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.

@@

Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used by~\cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

Each of those classes has a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle. The subject can browse the database by listening to those prototype sounds. The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed by~\cite{lafay2016JAES}.

@@

%La scène sonore est ainsi vue comme une somme de sources sonores, ou, si l'on admet la distinction opérée entre événements et textures sonores, «~un squelette d'événements sur un lit de textures~» \cite{nelken2013ear}. Cet outil de simulation est présenté en détail dans~\cite{rossignol2015simscene}.

As such, the underlying model of the scene is a sum of sound sources. The simulation interface is more toroughly described by~\cite{rossignol2015simscene}.

@@

The design of this experiment has been validated with a pilot study described by~\cite{lafay2014new}.

@@

\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni). To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}. With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:

@@

It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non ideal scenes. This observation is in agreement with the results obtained by~\cite{kuwano_memory_2003}. During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie~the events.

@@

We observe a noticeable difference in terms of class choices between the i- and non ideal scenes. The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes by~\cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non ideal scenes. These results confirm a fact previously observed by~\cite{raimbault2005urban,dubois2006cognitive}: the semantic nature of the sound sources play an important role in the assessment of the environment.

@@

Nevertheless, we notice some differences with results of~\cite{guastavino2006ideal} which show that sounds of \emph{public transportation} are specific of ideal urban soundscapes. The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors. Thus, in our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.

@@

This difference may be explained by the nature of the experimental protocol used in the two studies. As in our study, Guastavino asks the subjects to describe an environment. She asks them to perform this task starting only from their memories, whereas in our case subjects perform the same task using actual sound samples that they can listen to. The fact that subjects in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact. Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment, as studied by~\cite{lavandier2006contribution}.

@@

Considering $E$ and $T$ separately, it appears that: 1) the rank 5-precision obtained with $E$ is similar to the one obtained with $ET$ ; 2) the precision obtained with $T$ is always lower than the one obtained with $E$, by $10\%$ to $15\%$. Those results indicate that the semantic information that is discriminative is mostly carried by the events. Those results are in line with results of~\cite{maffiolo_caracterisation_1999}. As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie~by identifying the sources.

@@

The fact that the pleasantness of the ideal scenes is not correlated to global physical features, contrary to the pleasantness of non ideal scenes, has also been studied recently by~\cite{gozalo2015relationship}.

@@

This strongly confirms that ideal markers do have a positive impact on the perception of an environment. The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should be possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}. Those conclusions are in line with the positive approach introduced by~\cite{schafer1977tuning}.

@@

Considering sound scene synthesis or composition as part of an experimental protocol requires the availability of software resources that are easier and easier to master. While most processing framework that have been considered are developed in native language for efficiency purposes like Marsyas developed by~\cite{tzanetakis2000marsyas} or Clam developed by~\cite{amatriain2006clam}, many advanced sound manipulation.

Closer to our sound model is the Tapestrea framework proposed by~\cite{misra2007musical} that allows the manipulation sounds trough wavelet and sinusoidal modeling to allow sound manipulations for doing acousmatic composition or sound
design. The main issue with considering this framework for soundscape perception studies is that the sounds are processed using sound models that would strongly decrease the ecological validity of the study. That being said, the sequencing model is quite close to the one considered in this study.
